\documentclass{article}

\usepackage[margin=0.75in]{geometry}
\usepackage{qtree}
\qtreeprimes

\title{Syntax Trees in \LaTeX\ }
\date{}
\author{Rachael Steiner}

\begin{document}
\maketitle

This tutorial will show you how to make syntax trees using \LaTeX\ . It will assume you either know 
very basic \LaTeX\ (such as how to format a document) or can find out.

\section{Getting started}

To begin, you will need to use the package \textbf{qtree}. Include in your preamble:

\begin{verbatim}
	\usepackage{qtree}
\end{verbatim}

If you plan on using prime notation (e.g., for X-bar trees), also include the line:

\begin{verbatim}
	\qtreeprimes
\end{verbatim}

\section{The trees}

Start a tree with the command

\begin{verbatim}
	\Tree
\end{verbatim}

Trees are written in bracket notation. You can denote a node label by preceding 
it with a period. Any other text is taken to be the end of the branch.

\subsection{Simple Example}

Take, for example, the phrase ``the pretty tree'':

\begin{verbatim}
	\Tree [.NP [.Det the ] [.N [.Adj pretty ] [.N tree ] ] ]
\end{verbatim}

Looks like:

\Tree [.NP [.Det the ] [.N [.Adj pretty ] [.N tree ] ] ]

\newpage

\subsection{Formatting}

Doing the entire tree in one line can get pretty messy. Even that short phrase 
is a little cluttered. Luckily, \LaTeX\ ignores any extra whitespace in a tree,
so you can break it up across lines and indent it to make it a lot more readable.

Take the following example using ``a prettier tree'':

\begin{verbatim}
\Tree [.NP 
         [.Det a ] 
         [.N 
            [.Adj prettier ] 
            [.N tree ] 
         ] 
      ]
\end{verbatim}

Looks like:

\Tree [.NP 
			[.Det a ] 
			[.N 
				[.Adj prettier ] 
				[.N tree ] 
			] 
		]

\subsection{Bar notation}

To use X-bar notation, everything is more or less the same, but you use ``\textbackslash 1'' for the prime mark.

\begin{verbatim}
\Tree [.NP 
         [.DetP
            [.Det\1
               [.Det the ]
            ]
         ] 
         [.N\1 
            [.AdjP
               [.Adj\1
                  [.Adj prettiest ]
               ]
            ]
            [.N tree ] 
         ] 
      ]
\end{verbatim}

Looks like:

\Tree [.NP 
         [.DetP
            [.Det\1
               [.Det the ]
            ]
         ] 
         [.N\1 
            [.AdjP
               [.Adj\1
                  [.Adj prettiest ]
               ]
            ]
            [.N tree ] 
         ] 
      ]

\end{document}
